\documentclass[10pt]{article}

% Page layout — two-column via \twocolumn command
\usepackage[letterpaper,top=1in,bottom=1in,left=0.75in,right=0.75in]{geometry}

% Math
\usepackage{amsmath,amssymb}

% Tables
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{array}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=blue,urlcolor=blue,citecolor=blue}

% Typography
\usepackage{microtype}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{float}

% Figures
\usepackage{graphicx}
\usepackage{caption}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{tikz}
\pgfplotsset{compat=1.18}

% Code
\usepackage{listings}
\lstset{
  basicstyle=\ttfamily\fontsize{7}{8}\selectfont,
  breaklines=true,
  frame=single,
  backgroundcolor=\color{gray!10},
}

\captionsetup{font=small,labelfont=bf,skip=4pt}
\setlength{\columnsep}{0.28in}
\setlength{\parskip}{2pt}
\setlength{\parindent}{1em}

% Tighter list spacing globally
\setlist{noitemsep,topsep=2pt,leftmargin=*}

\begin{document}

% ---- Full-width header (title + authors + abstract) ----
\twocolumn[{%
\begin{center}
\vspace*{0.15in}
{\Large\bfseries Improving Multi-Constraint Instruction Following in\\[3pt]
1B-Parameter Language Models via RECAST Fine-Tuning\par}
\vspace{9pt}
{\normalsize
  Shreyas Natarajan \quad
  Fnu Rinkle Rose Renny \quad
  Mark Aashish Lnu \quad
  Gayathri Ethiraj\par}
\vspace{4pt}
{\small\textit{Masters in Data Science, University of Wisconsin--Madison}\par}
{\small\texttt{\{natarajan9, rinklerosere, mlnu5, gethiraj\}@wisc.edu}\par}
{\small STAT 453, Spring 2026 $\cdot$ Team 15\par}
\vspace{10pt}
{\bf Abstract}
\end{center}
\begin{quotation}\small
Small language models ($\sim$1B parameters) offer practical advantages in
cost and inference speed yet consistently underperform on multi-constraint
instruction-following tasks. We present an end-to-end pipeline---9-stage
data preprocessing and augmentation (30K $\to$ $\sim$90K examples),
supervised fine-tuning via full parameter updates and LoRA, hybrid
rule-based and LLM-judge evaluation, and 5-fold cross-validation---to
quantify and improve constraint-following in 1B-parameter models using the
RECAST-30K dataset~\cite{guo2025recast}. Evaluating Llama 3.2 1B Instruct,
the zero-shot baseline achieves a Hard Constraint Satisfaction Rate (CSR)
of 28.9\%. LoRA fine-tuning (rank 8, targeting \texttt{q\_proj} and
\texttt{v\_proj}) reaches a 5-fold mean CSR of \textbf{80.0\%},
substantially outperforming full fine-tuning (33.2\% $\pm$ 2.3\%). Length
adherence (39.2\%) and keyword inclusion (72.7\%) remain the hardest
constraints; structural and qualitative categories (format, tone, style)
reach near-perfect pass rates after LoRA training.
\end{quotation}
\vspace{10pt}
}]

% ----------------------------------------------------------------
\section{Introduction}
% ----------------------------------------------------------------

Large language models are powerful but expensive to deploy at scale.
Models in the 1--2B parameter range offer substantially lower compute cost
and faster inference latency, yet they struggle with
\textit{constraint-following}: simultaneously satisfying explicit
requirements on format, length, tone, style, keywords, and structure
imposed by a prompt. This difficulty compounds as the number of
simultaneous constraints grows---a setting that directly mirrors
real-world scenarios such as drafting structured reports, composing
constrained marketing copy, or generating formatted API responses.

Benchmarks such as IFEval~\cite{zhou2023ifeval} and
FollowBench~\cite{jiang2024followbench} have documented these failures
comprehensively. As a reference point, Llama-3.2-3B achieves only
$\sim$77.4\% prompt-strict accuracy on IFEval; the 1B variant is weaker
still. The RECAST framework~\cite{guo2025recast} addresses the data side
by synthesizing multi-constraint training sets with automated validators
enabling fine-grained, constraint-level supervision. RECAST fine-tuning at
7--8B scale outperforms much larger instruct-tuned models, but behavior at
1B scale is unexplored. Separately, LoRA~\cite{hu2022lora} and full
fine-tuning have been studied for reasoning and code tasks, but not
specifically for constraint-following~\cite{shuttleworth2024lora}.

\textbf{Research questions.} This project addresses three open questions:
\begin{enumerate}
  \item Can RECAST-based fine-tuning meaningfully improve
        constraint-following in 1B-parameter models?
  \item Does full fine-tuning outperform LoRA, or do low-rank updates
        suffice?
  \item Which constraint categories benefit most, and which remain
        difficult?
\end{enumerate}

% ----------------------------------------------------------------
\section{Related Work}
% ----------------------------------------------------------------

\textbf{Constraint-following benchmarks.}
IFEval~\cite{zhou2023ifeval} introduced a structured evaluation suite
covering 25 verifiable instruction categories. FollowBench~\cite{jiang2024followbench}
extended this with multi-level difficulty constraints, showing sharp
degradation as constraint count increases. Both benchmarks diagnose the
problem but do not provide training recipes.

\textbf{RECAST.}
Guo et al.~\cite{guo2025recast} propose a constraint back-translation
framework that synthesizes multi-constraint training data by extracting
constraints from existing high-quality responses and injecting them back
into prompts, guaranteeing physical achievability. Every constraint has an
automated validator, enabling fine-grained training signal. At 7B--8B
scales, RECAST outperforms larger instruct-tuned models; behavior at 1B
parameters is unexplored.

\textbf{LoRA vs.\ full fine-tuning.}
Hu et al.~\cite{hu2022lora} introduced Low-Rank Adaptation, which freezes
the base model and injects trainable rank-$r$ matrices into attention
projections, training only $\sim$0.5\% of parameters. Shuttleworth et
al.~\cite{shuttleworth2024lora} show that full fine-tuning adjusts existing
singular vectors more conservatively while LoRA introduces ``intruder
dimensions'' that can accelerate catastrophic forgetting---but this
analysis targets reasoning and code tasks. Whether these dynamics hold for
constraint-following remains open. Table~\ref{tab:lora_vs_ft} summarizes
the practical tradeoffs.

\begin{table}[H]
\centering
\caption{LoRA vs.\ full fine-tuning at 1B scale. Both methods are
novel for the constraint-following setting.}
\label{tab:lora_vs_ft}
\small
\begin{tabular}{lcc}
\toprule
\textbf{Dimension} & \textbf{LoRA} & \textbf{Full FT} \\
\midrule
Trainable params  & $\sim$0.5\%      & 100\%           \\
VRAM peak         & $\sim$7--9\,GB   & $\sim$16--20\,GB \\
Storage/run       & $\sim$30--60\,MB & $\sim$2.4\,GB   \\
Forgetting risk   & Lower            & Higher          \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Instruction-tuning data.}
The Tulu corpus~\cite{ivison2023tulu} assembles diverse
instruction-following examples from FLAN, ShareGPT, OpenAssistant, and
other sources, providing broad coverage needed to prevent catastrophic
forgetting when fine-tuning on the narrower RECAST constraint distribution.

% ----------------------------------------------------------------
\section{Dataset and Preprocessing}
% ----------------------------------------------------------------

\subsection{RECAST-30K}
The RECAST-30K dataset provides multi-constraint instruction--response
pairs covering 15 constraint categories (Style, Topic, Length, Keyword,
End\_With, Start\_With, Language, Format, Tone, Role-Playing, Numerical,
etc.) with an average of $\sim$13.4 constraints per instruction and 90.1\%
human annotator agreement. Each record contains:
\begin{itemize}
  \item \texttt{winner\_prompt} --- the multi-constraint instruction;
  \item \texttt{response\_of\_winner\_prompt} --- a constraint-satisfying
        gold response;
  \item \texttt{added\_constraint} --- dict mapping category to
        constraint descriptions;
  \item \texttt{rule\_evaluate\_dict} --- structured validator parameters
        (word length bounds, required keywords, start/end strings).
\end{itemize}

\subsection{Preprocessing Pipeline}
The raw JSONL passes through a nine-stage pipeline:
\begin{enumerate}[label=\arabic*.]
  \item \textbf{Language detection.} Non-English prompts filtered via
        \texttt{langdetect} (ISO-639-1 check on first 2000 chars).
  \item \textbf{HTML/entity removal.} Tags and entities stripped by
        compiled regex.
  \item \textbf{Emoji/control removal.} Emojis replaced with spaces;
        invisible Unicode and long symbol runs removed.
  \item \textbf{Quality gates.} Prompts $<$15 words, $<$85\% printable,
        or with empty responses dropped.
  \item \textbf{Constraint integrity.} Validates \texttt{added\_constraint}
        structure and count consistency.
  \item \textbf{Stopword removal.} Applied to prompts only, after quality
        gates.
  \item \textbf{Exact deduplication.} SHA-256 fingerprints on normalized
        prompts.
  \item \textbf{Near-deduplication.} MinHash LSH (\texttt{datasketch},
        Jaccard $\tau{=}0.85$, 128 permutations).
  \item \textbf{Augmentation.} Two techniques expand cleaned $\sim$30K
        records to $\sim$90K: \textit{lexical editing} via synonym
        replacement ($+{\sim}30$K) and \textit{back-translation} via
        English$\to$German$\to$English round-trip ($+{\sim}30$K).
\end{enumerate}
A post-cleaning distribution audit flags categories below 50\% of the
expected uniform share. The pipeline runs in parallel with 8 workers.

\subsection{Data Splits}
For cross-validation, the $\sim$90K augmented dataset is partitioned into
5 folds with a 72\% / 18\% / 10\% train/validation/test split per fold.

% ----------------------------------------------------------------
\section{Methods}
% ----------------------------------------------------------------

\subsection{Full Fine-Tuning}
The full fine-tuning pipeline makes all $\sim$1.24B model parameters
trainable using HuggingFace \texttt{Trainer} with gradient checkpointing.
Each record is formatted as a user/assistant chat turn via the Llama-3
chat template; prompt tokens are masked (\texttt{labels}$=-100$) so loss
is computed only on the response.

\textbf{Loss.} Standard cross-entropy over response tokens. An earlier
custom penalty $(1 - s/\text{total})$ encountered backpropagation
difficulties; CE loss provided stable gradient-based optimization and is a
principled probabilistic training objective under teacher forcing.

\textbf{Hyperparameters:} AdamW optimizer; lr $= 2{\times}10^{-5}$;
cosine LR scheduler with 3\% warmup; effective batch 32 (per-device batch
4, grad accum 8); 2 epochs; max 512 tokens; bfloat16 on A100 40\,GB;
gradient checkpointing enabled.

\subsection{LoRA Fine-Tuning}
The LoRA pipeline applies rank-$r$ adapter matrices to \texttt{q\_proj}
and \texttt{v\_proj} in all attention layers:
\begin{equation}
  W' = W + \frac{\alpha}{r} \cdot BA
  \label{eq:lora}
\end{equation}
where $W \in \mathbb{R}^{d \times d}$ is frozen, $A \in \mathbb{R}^{r
\times d}$ and $B \in \mathbb{R}^{d \times r}$ are trainable, $r{=}8$,
$\alpha{=}16$, dropout $= 0.05$. This trains 81,500 adapter parameters
($\approx$0.007\% of weights); the adapter ($\sim$30--60\,MB) is saved
while the base model remains frozen.

The pipeline uses \texttt{trl.SFTTrainer} with the Llama-3 chat format:

\begin{lstlisting}
<|begin_of_text|><|start_header_id|>user<|end_header_id|>
{instruction}<|eot_id|>
<|start_header_id|>assistant<|end_header_id|>
{response}<|eot_id|>
\end{lstlisting}

\textbf{Hyperparameters:} lr $= 10^{-4}$; cosine scheduler; 10 warmup
steps; weight decay 0.01; effective batch 32 (batch 8, grad accum 4); 1
epoch; max 512 tokens; bfloat16.

\subsection{Evaluation: Constraint Validation}

\textbf{Rule-based mode.} Four constraints are evaluated deterministically
from \texttt{rule\_evaluate\_dict}:

\begin{table}[H]
\centering
\caption{Rule-based constraint checks. Each returns 0 (fail), 1 (pass),
or \texttt{None} (not applicable).}
\label{tab:rule_checks}
\small
\begin{tabular}{ll}
\toprule
\textbf{Constraint} & \textbf{Check} \\
\midrule
\texttt{length}      & Word count in $[\min, \max]$ \\
\texttt{keyword}     & Each keyword appears $\geq$ required count \\
\texttt{start\_with} & Response begins with required string \\
\texttt{end\_with}   & Response ends with required string \\
\bottomrule
\end{tabular}
\end{table}

\textbf{LLM-judge mode.} For qualitative constraints (style, tone,
role-playing, numerical), a judge model (default: Qwen2.5-7B-Instruct;
also Gemma-2-9B-Instruct at 4-bit) receives the response plus
plain-English constraint descriptions and outputs a JSON verdict.

\textbf{Scoring metrics:}
\begin{align}
  \text{CSR}^{(i)} &= \frac{\text{constraints satisfied in example } i}
                          {\text{constraints checked in example } i} \\
  \text{Hard CSR}  &= \mathbb{1}[\text{all checked constraints satisfied}]
\end{align}
The standalone \texttt{ConstraintChecker} class dispatches across 18
constraint types in 8 categories: length, keywords, structural,
capitalization, format, language, punctuation, and detectable.

\subsection{5-Fold Cross-Validation}
The $\sim$90K augmented dataset is partitioned into 5 folds
(72\% / 18\% / 10\% train/val/test). The model is trained on $k{-}1$
folds and evaluated on the held-out fold, providing variance estimates
for both LoRA and full fine-tuning under the same framework.

% ----------------------------------------------------------------
\section{Experimental Setup}
% ----------------------------------------------------------------

\textbf{Hardware.} All training ran on Google Colab Pro+ with NVIDIA A100
40\,GB GPUs. Models were loaded in bfloat16.

\textbf{Models.} Primary model: \textbf{Llama 3.2 1B Instruct} (1.24B
parameters, publicly available instruction-tuned causal decoder-only
Transformer). Secondary (exploratory runs): Qwen 2.5 1.5B Instruct.

\begin{table}[H]
\centering
\caption{Hardware requirements per experimental configuration.}
\label{tab:compute}
\small
\begin{tabular}{lccc}
\toprule
\textbf{Run type} & \textbf{VRAM} & \textbf{Time} & \textbf{Params} \\
\midrule
LoRA (r=8)   & $\sim$7--9\,GB   & 2--4\,hr    & 81,500 \\
Full FT       & $\sim$16--20\,GB & 4.5--8\,hr  & 1.24B  \\
LLM judge (7B)& $\sim$14\,GB    & 30\,min/500 & ---    \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Evaluation protocol:} (1) zero-shot baseline on RECAST test split;
(2) LoRA: merge adapter with base, evaluate on held-out fold per CV round;
(3) Full FT: load checkpoint, evaluate on same fold; (4) per-category
analysis across all training regimes.

% ----------------------------------------------------------------
\section{Results}
% ----------------------------------------------------------------

\subsection{Zero-Shot Baseline}

The zero-shot Llama 3.2 1B Instruct baseline achieves an overall Hard CSR
of \textbf{28.9\%} on the RECAST test set. Per-constraint pass rates
(Table~\ref{tab:baseline}, Figure~\ref{fig:baseline_bar}) reveal a
strongly bimodal pattern: structural constraints (sentence length, format,
capitalization, punctuation) are already satisfied near-perfectly by the
pretrained model, while content-dependent constraints fail frequently.
Hard CSR is low because even a single failure among the $\sim$13.4 average
constraints per example counts as complete failure.

\begin{table}[H]
\centering
\caption{Zero-shot baseline per-constraint pass rates for Llama 3.2 1B
Instruct.}
\label{tab:baseline}
\small
\begin{tabular}{lcc}
\toprule
\textbf{Constraint} & \textbf{Pass rate} & \textbf{Notes} \\
\midrule
\texttt{word\_length}      & 32.0\% & Hardest rule-based \\
\texttt{keyword}           & 40.0\% & Requires explicit counting \\
\texttt{start\_with}       & 56.5\% & Partially structural \\
\texttt{end\_with}         & 36.0\% & Hard to control endpoint \\
\texttt{sentence\_length}  & 100.0\% & Trivially satisfied \\
\texttt{format}            & 100.0\% & Follows conventions \\
\texttt{all\_caps/lower}   & 100.0\% & Easy to learn \\
\texttt{no\_commas}        & 100.0\% & Rare; default satisfied \\
\midrule
\textbf{Hard CSR}          & \textbf{28.9\%} & All must pass \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[htbp]
\centering
\begin{tikzpicture}
\begin{axis}[
  ybar,
  bar width=9pt,
  width=\columnwidth,
  height=5.2cm,
  ylabel={Pass Rate},
  ymin=0, ymax=1.25,
  ytick={0,0.2,0.4,0.6,0.8,1.0},
  yticklabel={\pgfmathprintnumber[fixed,precision=1]{\tick}},
  xtick=data,
  xticklabels={
    word\_len, keyword, start, end,
    sent\_len, format, caps, commas
  },
  x tick label style={rotate=35,anchor=east,font=\scriptsize},
  nodes near coords,
  nodes near coords style={font=\tiny,
    /pgf/number format/fixed,
    /pgf/number format/precision=2},
  enlarge x limits=0.1,
  title={Baseline Zero-Shot Per-Constraint Pass Rates},
  title style={font=\scriptsize},
]
\addplot[fill=gray!60] coordinates {
  (1,0.320)(2,0.400)(3,0.565)(4,0.360)
  (5,1.000)(6,1.000)(7,1.000)(8,1.000)
};
\end{axis}
\end{tikzpicture}
\caption{Zero-shot baseline pass rates. Content-dependent constraints
(word count, keyword, start/end) score 32--57\%; structural constraints
are trivially satisfied at baseline.}
\label{fig:baseline_bar}
\end{figure}

\subsection{LoRA Fine-Tuning Results (5-Fold CV)}

LoRA fine-tuning achieves a 5-fold mean CSR of \textbf{80.0\%} with very
low fold-to-fold variance (Table~\ref{tab:lora_folds}), representing a
$+$51.1 pp improvement over the zero-shot baseline.

\begin{table}[H]
\centering
\caption{LoRA 5-fold CSR results (Llama 3.2 1B, $r{=}8$).}
\label{tab:lora_folds}
\small
\begin{tabular}{ccc}
\toprule
\textbf{Fold} & \textbf{CSR} & \textbf{N} \\
\midrule
1 & 0.7972 & 200 \\
2 & 0.7972 & 200 \\
3 & 0.7967 & 200 \\
4 & 0.8039 & 200 \\
5 & 0.8050 & 200 \\
\midrule
\textbf{Mean} & \textbf{0.8000} & 1000 \\
\bottomrule
\end{tabular}
\end{table}

Per-constraint pass rates (Table~\ref{tab:lora_per_constraint} and
Figure~\ref{fig:lora_bar}) reveal a clear bifurcation: structural and
qualitative constraints reach ceiling performance ($\geq$99.7\%), while
length adherence (39.2\%) and keyword inclusion (72.7\%) remain
challenging. Figure~\ref{fig:lora_folds} confirms fold-to-fold variance is
$<1\%$, indicating stable generalization independent of the train/test
split.

\begin{table}[H]
\centering
\caption{LoRA per-constraint pass rates (mean $\pm$ std across 5 folds).}
\label{tab:lora_per_constraint}
\small
\begin{tabular}{lcc}
\toprule
\textbf{Constraint} & \textbf{Mean} & \textbf{Std} \\
\midrule
\texttt{length}        & 0.3920 & 0.0244 \\
\texttt{keyword}       & 0.7270 & 0.0351 \\
\texttt{start\_with}   & 0.5830 & 0.0413 \\
\texttt{end\_with}     & 0.5020 & 0.0182 \\
\texttt{format}        & 0.9970 & 0.0045 \\
\texttt{tone}          & 0.9990 & 0.0022 \\
\texttt{style}         & 1.0000 & 0.0000 \\
\texttt{role\_playing} & 1.0000 & 0.0000 \\
\texttt{numerical}     & 1.0000 & 0.0000 \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure*}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  ybar,
  bar width=16pt,
  width=0.82\textwidth,
  height=6cm,
  ylabel={Pass Rate (mean $\pm$ std)},
  ymin=0, ymax=1.2,
  ytick={0,0.2,0.4,0.6,0.8,1.0},
  xtick=data,
  xticklabels={
    Length, Keyword, Start\_With, End\_With,
    Format, Tone, Style, RolePlay, Numerical
  },
  x tick label style={rotate=30,anchor=east,font=\small},
  nodes near coords,
  nodes near coords style={font=\scriptsize,
    /pgf/number format/fixed,
    /pgf/number format/precision=2},
  enlarge x limits=0.06,
  error bars/y dir=both,
  error bars/y explicit,
  title={LoRA Per-Constraint CSR --- Mean $\pm$ Std Across 5 Folds
    ($r{=}8$, lr$=10^{-4}$, Llama-3.2-1B)},
  title style={font=\small},
]
\addplot+[
  fill=blue!60,
  error bars/.cd, y dir=both, y explicit
] coordinates {
  (1,0.3920)+-(0,0.0244)
  (2,0.7270)+-(0,0.0351)
  (3,0.5830)+-(0,0.0413)
  (4,0.5020)+-(0,0.0182)
  (5,0.9970)+-(0,0.0045)
  (6,0.9990)+-(0,0.0022)
  (7,1.0000)+-(0,0.0000)
  (8,1.0000)+-(0,0.0000)
  (9,1.0000)+-(0,0.0000)
};
\end{axis}
\end{tikzpicture}
\caption{LoRA per-constraint pass rates with $\pm$std error bars across 5
folds. Structural and qualitative constraints (format, tone, style,
role-playing, numerical) reach near-perfect performance; length adherence
(39.2\%) remains the largest challenge.}
\label{fig:lora_bar}
\end{figure*}

\begin{figure}[htbp]
\centering
\begin{tikzpicture}
\begin{axis}[
  ybar,
  bar width=18pt,
  width=\columnwidth,
  height=5cm,
  ylabel={CSR (\%)},
  xlabel={Fold},
  ymin=70, ymax=85,
  ytick={72,74,76,78,80,82,84},
  xtick=data,
  xticklabels={1,2,3,4,5},
  nodes near coords,
  nodes near coords style={font=\scriptsize,
    /pgf/number format/fixed,
    /pgf/number format/precision=1},
  enlarge x limits=0.3,
  title={LoRA CSR per Fold},
  title style={font=\small},
]
\addplot[fill=blue!50] coordinates {
  (1,79.72)(2,79.72)(3,79.67)(4,80.39)(5,80.50)
};
\addplot[red,dashed,domain=0.5:5.5,samples=2]{80.0};
\legend{CSR per fold, Mean $= 80.0\%$}
\end{axis}
\end{tikzpicture}
\caption{LoRA CSR per fold. Dashed red line marks the 5-fold mean
(80.0\%). Variance is $<1\%$, confirming stable generalization.}
\label{fig:lora_folds}
\end{figure}

\subsection{Full Fine-Tuning Results (5-Fold CV)}

Full fine-tuning of all 1.24B parameters achieves a 5-fold mean CSR of
\textbf{33.2\%} ($\pm$2.3\%), substantially below LoRA
(Table~\ref{tab:ft_folds}). The abnormally low train loss in Fold~1
($=0$) likely reflects a logging artifact. Eval loss is stable across
folds ($\pm$0.0024); the CSR range of 30.4--36.6\% indicates greater
sensitivity to data partitioning than LoRA.

\begin{table}[H]
\centering
\caption{Full fine-tuning 5-fold results on Llama 3.2 1B.}
\label{tab:ft_folds}
\small
\begin{tabular}{lrrr}
\toprule
\textbf{Fold} & \textbf{Eval loss} & \textbf{Train loss} & \textbf{CSR} \\
\midrule
1 & 1.4217 & \phantom{0}0.0000 & 36.6\% \\
2 & 1.4215 & 10.9864           & 30.4\% \\
3 & 1.4177 & 10.9924           & 33.6\% \\
4 & 1.4251 & 10.9826           & 31.0\% \\
5 & 1.4229 & 10.9752           & 34.6\% \\
\midrule
\textbf{Mean} & 1.4218 & 8.7873 & \textbf{33.2\%} \\
\textbf{Std}  & 0.0024 & 4.3937 & 2.3\% \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[htbp]
\centering
\begin{tikzpicture}
\begin{axis}[
  ybar,
  bar width=18pt,
  width=\columnwidth,
  height=5cm,
  ylabel={CSR (\%)},
  xlabel={Fold},
  ymin=0, ymax=50,
  xtick=data,
  xticklabels={1,2,3,4,5},
  nodes near coords,
  nodes near coords style={font=\scriptsize,
    /pgf/number format/fixed,
    /pgf/number format/precision=1},
  enlarge x limits=0.3,
  title={Full Fine-Tuning CSR per Fold},
  title style={font=\small},
]
\addplot[fill=green!55] coordinates {
  (1,36.6)(2,30.4)(3,33.6)(4,31.0)(5,34.6)
};
\addplot[red,dashed,domain=0.5:5.5,samples=2]{33.2};
\legend{CSR per fold, Mean $= 33.2\%$}
\end{axis}
\end{tikzpicture}
\caption{Full fine-tuning CSR per fold with mean line (dashed). Fold~1
shows the best performance (36.6\%); variance is higher than LoRA.}
\label{fig:ft_folds}
\end{figure}

\subsection{Head-to-Head Comparison}

Table~\ref{tab:comparison} and Figure~\ref{fig:comparison_bar} summarize
results across all three training regimes. LoRA delivers a $+$51.1 pp gain
over the zero-shot baseline and a $+$46.8 pp gain over full fine-tuning.

\begin{table}[H]
\centering
\caption{Three-way model comparison. PR = per-constraint pass rate.
Full FT per-constraint breakdown was not captured at evaluation time.}
\label{tab:comparison}
\small
\begin{tabular}{lcccc}
\toprule
\textbf{Config} & \textbf{CSR} & \textbf{Len PR} & \textbf{Kw PR}
  & \textbf{Start PR} \\
\midrule
Baseline         & 28.9\%         & 32.0\% & 40.0\% & 56.5\% \\
Full FT          & 33.2\%         & ---    & ---    & ---    \\
LoRA ($r{=}8$)   & \textbf{80.0\%}& 39.2\% & 72.7\% & 58.3\% \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[htbp]
\centering
\begin{tikzpicture}
\begin{axis}[
  ybar,
  bar width=26pt,
  width=\columnwidth,
  height=5.5cm,
  ylabel={Overall CSR (\%)},
  ymin=0, ymax=100,
  xtick=data,
  xticklabels={Baseline, Full FT, LoRA ($r{=}8$)},
  x tick label style={font=\small},
  nodes near coords,
  nodes near coords style={font=\normalsize\bfseries},
  enlarge x limits=0.3,
  title={Overall CSR: Baseline vs.\ Full FT vs.\ LoRA},
  title style={font=\small},
]
\addplot[fill=gray!60]  coordinates{(1,28.9)};
\addplot[fill=green!60] coordinates{(2,33.2)};
\addplot[fill=blue!60]  coordinates{(3,80.0)};
\end{axis}
\end{tikzpicture}
\caption{Overall CSR comparison. LoRA substantially outperforms both
the zero-shot baseline ($+$51.1 pp) and full fine-tuning ($+$46.8 pp).}
\label{fig:comparison_bar}
\end{figure}

% ----------------------------------------------------------------
\section{Discussion}
% ----------------------------------------------------------------

\subsection{Key Findings}

\textbf{LoRA outperforms full fine-tuning by a large margin.}
The 80.0\% vs.\ 33.2\% gap ($+$46.8 pp) is the most striking result and
contradicts the hypothesis that full parameter updates should give the
optimizer more expressive power. We attribute the reversal to
\textit{catastrophic forgetting}: updating all 1.24B parameters on the
narrow RECAST constraint distribution tends to overwrite the broader
instruction-following competence built into the pretrained checkpoint.
LoRA's frozen base model preserves this competence while adapting attention
projections to constraint-following patterns.

\textbf{Length adherence is the hardest constraint.}
Despite a $+$7.2 pp gain (baseline 32.0\% $\to$ LoRA 39.2\%), word-count
adherence remains below 40\%. The model must maintain an accurate running
count of generated tokens while simultaneously satisfying other constraints
and producing coherent text---a form of explicit bookkeeping that is
difficult to learn from sequence-level supervision alone.

\textbf{Structural and qualitative constraints saturate.}
Format, tone, style, role-playing, and numerical constraints all reach
$\geq$99.7\% pass rates under LoRA. These require a stylistic shift in the
response distribution rather than explicit counting, and tend to be
well-represented in the training data.

\textbf{Loss function design.}
The custom penalty $(1-s/\text{total})$, while conceptually intuitive,
encountered gradient computation issues during backpropagation. Standard
cross-entropy over constraint-satisfying gold responses proved more
effective, as CE is a differentiable probabilistic surrogate well-suited to
the teacher-forcing paradigm.

\subsection{Design Decisions}

\textbf{Augmentation to $\sim$90K.} Tripling the training set via lexical
editing and back-translation helps LoRA generalize beyond exact phrasings
while preserving constraint semantics.

\textbf{5-fold CV.} The low fold-to-fold LoRA variance (CSR range:
79.67--80.50\%) demonstrates robustness. The higher variance for full FT
(30.4--36.6\%) suggests full FT is more sensitive to training partition.

\textbf{Prompt masking.} Masking prompt tokens ensures CE loss is computed
only over the response, preventing gradient dilution from the long
constraint specifications embedded in RECAST prompts.

\subsection{Limitations}
\begin{enumerate}
  \item \textbf{Model scale.} Resource constraints limited primary
        experiments to Llama 3.2 1B; the proposal targeted four 1--2B
        models. Conclusions about scale sensitivity are preliminary.
  \item \textbf{Compute cost.} Full FT required $\sim$4.5 hr/fold on A100,
        limiting hyperparameter exploration.
  \item \textbf{Evaluation coverage.} The deterministic evaluator covers
        only 4 of 15+ constraint categories; the LLM judge adds inference
        cost and may introduce systematic bias (7B judge evaluating 1B
        outputs may differ from human annotators).
  \item \textbf{Catastrophic forgetting.} We did not evaluate on IFEval
        after full fine-tuning; direct measurement of general
        instruction-following degradation remains future work.
  \item \textbf{Full FT per-constraint breakdown.} Per-constraint pass
        rates were not captured for full FT, preventing direct
        per-category comparison between methods.
\end{enumerate}

% ----------------------------------------------------------------
\section{Conclusion}
% ----------------------------------------------------------------

We built and evaluated a complete end-to-end pipeline for multi-constraint
instruction-following in 1B-parameter language models, covering 9-stage
data preprocessing with augmentation (30K $\to$ 90K), two fine-tuning
regimes (full FT and LoRA $r{=}8$), hybrid evaluation (rule-based $+$ LLM
judge over 9 constraint types), and 5-fold cross-validation.

The zero-shot baseline achieves Hard CSR 28.9\%, with length (32.0\%) and
keyword (40.0\%) as the weakest categories. LoRA raises overall CSR to
80.0\% (5-fold mean), with structural and qualitative constraints near
ceiling. Full fine-tuning achieves only 33.2\%, likely due to catastrophic
forgetting of the pretrained instruction-following competence.

The central finding: \textbf{LoRA substantially outperforms full
fine-tuning} for constraint-following at 1B scale. Training 0.007\% of
parameters outperforms training 100\%, using half the VRAM, a fraction of
the training time, and producing a $40{\times}$ smaller checkpoint. This
validates LoRA as the preferred adaptation strategy and motivates future
investigation into higher ranks ($r{=}16$, $r{=}32$), targeted
length-heavy augmentation, and multi-model robustness on Gemma 3 1B and
SmolLM2 1.7B.

% ----------------------------------------------------------------
\section*{Team Contributions}
% ----------------------------------------------------------------

\textbf{Shreyas Natarajan:} Dataset preprocessing pipeline design and
implementation (stages 1--8); constraint integrity validation; pipeline
parallelization; overall system integration and evaluation framework.

\textbf{Fnu Rinkle Rose Renny:} Back-translation augmentation
(\texttt{back\_translate.py}); near-deduplication with MinHash LSH; LoRA
fine-tuning pipeline; cross-validation framework for LoRA; result
analysis and visualization.

\textbf{Mark Aashish Lnu:} Lexical editing augmentation
(\texttt{lexical\_edit.py}); full fine-tuning pipeline
(\texttt{train\_full\_finetune.py}); 5-fold cross-validation framework;
evaluation metrics (CSR, Hard CSR); LLM-judge integration.

\textbf{Gayathri Ethiraj:} Dataset clustering analysis;
\texttt{ConstraintChecker} implementation; evaluation visualization
(\texttt{viz\_utils.py}); per-constraint analysis and result
interpretation; report writing and literature review.

% ----------------------------------------------------------------
\begin{thebibliography}{9}

\bibitem{zhou2023ifeval}
J.~Zhou et al.,
``Instruction-Following Evaluation for Large Language Models,''
\textit{arXiv}:2311.07911, 2023.

\bibitem{jiang2024followbench}
Y.~Jiang et al.,
``FollowBench: A Multi-level Fine-grained Constraints Following Benchmark
for Large Language Models,''
\textit{Proc.\ ACL 2024}, pp.\ 4667--4688.

\bibitem{guo2025recast}
Z.~Guo et al.,
``RECAST: Expanding the Boundaries of LLMs' Complex Instruction Following
with Multi-Constraint Data,''
\textit{arXiv}:2505.19030, 2025.

\bibitem{hu2022lora}
E.~J.~Hu et al.,
``LoRA: Low-Rank Adaptation of Large Language Models,''
\textit{ICLR 2022}. \textit{arXiv}:2106.09685.

\bibitem{shuttleworth2024lora}
R.~Shuttleworth et al.,
``LoRA vs Full Fine-tuning: An Illusion of Equivalence,''
\textit{arXiv}:2410.21228, 2024.

\bibitem{ivison2023tulu}
H.~Ivison et al.,
``Camels in a Changing Climate: Enhancing LM Adaptation with Tulu 2,''
\textit{arXiv}:2311.10702, 2023.

\bibitem{hoffmann2022chinchilla}
J.~Hoffmann et al.,
``Training Compute-Optimal Large Language Models,''
\textit{NeurIPS 2022}.

\end{thebibliography}

% ----------------------------------------------------------------
\appendix
\section*{Appendix: AI Assistance Disclosure}
% ----------------------------------------------------------------

Per course policy, this report was written with the assistance of Claude
Sonnet 4.6 (Anthropic, 2026) for LaTeX formatting, prose editing, and
structural organization. Prompts instructed the model to reformat the
existing content into a two-column conference paper style (NeurIPS/ICML
template), condense verbose descriptions while preserving all experimental
results, and add team contributions. All experimental results, numbers,
figures, and scientific conclusions were produced by the team and not
generated by AI.

\end{document}
